Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-28 paper.tex
Shuffle Jeff and Eric's affils

%%% Use the \authfn to add symbols for additional footnotes, if any.
% 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
% First author
\author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
\author[4,\authfn{1}]{Will Tyler}
\author[5,\authfn{1}]{Tom White}

@@

% Senior
\author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
\author[10,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253

\affil[1]{Related Sciences}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Independent researcher}
\affil[5]{Tom White Consulting Ltd.}
\affil[6]{Our Future Health, Manchester, UK.}
\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
Hospital, Boston, Massachusetts 02114, USA}
\affil[9]{NVIDIA Ltd, Reading, UK}
\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
University of Oxford, UK}
\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
\affil[12]{Wellcome Sanger Institute}
\affil[13]{Genomics England}
\affil[14]{Department of Cell and Molecular Biology, National
  Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
  Uppsala University, Uppsala, Sweden}
\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}

%%% Author Notes
\authnote{\authfn{1}Joint first author.}
\authnote{\authfn{2}Joint senior author.}
\authnote{\authfn{3}jerome.kelleher@bdi.ox.ac.uk}

@@

\subsection{Future applications: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
we uploaded the Zarr directory hierarchy for the Genomics England
data chromosome 2 described in Table~\ref{tab-genomics-england-data}
to an AWS S3 standard bucket. 
We repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

@@

\subsection{AWS parallel Zarr prototype}
[TODO Implementation of prototype Zarr implementaion]

>>>>

%%% Use the \authfn to add symbols for additional footnotes, if any.
% 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
% First author
\author[1,\authfn{1},\authfn{3}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
\author[4,\authfn{1}]{Will Tyler}
\author[5,\authfn{1}]{Tom White}

@@

% Senior
\author[1,\authfn{2},\authfn{3}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
\author[10,\authfn{2},\authfn{4}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253

\affil[1]{Open Athena AI Foundation}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Independent researcher}
\affil[5]{Tom White Consulting Ltd.}
\affil[6]{Our Future Health, Manchester, UK.}
\affil[7]{Program in Medical and Population Genetics, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
\affil[8]{Analytic and Translational Genetics Unit, Massachusetts General
Hospital, Boston, Massachusetts 02114, USA}
\affil[9]{NVIDIA Ltd, Reading, UK}
\affil[10]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
University of Oxford, UK}
\affil[11]{Novo Nordisk Foundation Center for Genomic Mechanisms of Disease, Broad Institute of MIT and Harvard, Cambridge, Massachusetts 02142, USA}
\affil[12]{Wellcome Sanger Institute}
\affil[13]{Genomics England}
\affil[14]{Department of Cell and Molecular Biology, National
  Bioinformatics Infrastructure Sweden, Science for Life Laboratory,
  Uppsala University, Uppsala, Sweden}
\affil[15]{School of Computer Science, McGill University, Montreal, QC, Canada}

%%% Author Notes
\authnote{\authfn{1}Joint first author.}
\authnote{\authfn{2}Joint senior author.}
\authnote{\authfn{3}Work partially performed at Related Sciences.}
\authnote{\authfn{4}jerome.kelleher@bdi.ox.ac.uk}

@@

\subsection{Future applications: Cloud computing}
To explore the use of cloud computing services with VCF Zarr,
Genomics England provided access to data within their secure
cloud environment. The VCF Zarr files for Chromosome 2
described in Table~\ref{tab-genomics-england-data}
were hosted in a standard AWS S3 bucket.

Using this data we repeated the benchmarks above on the same subset
of 562,640 variants and 78,195 samples using the same
single-threaded sequential Zarr code.
Extracting the position array required 15s CPU, 23.8s elapsed.
The af-dist calculation required 2.70min CPU, 5.23min elapsed.
Creating the filter mask required 3.05min CPU, 11.15min elapsed.

@@

\subsection{AWS parallel Zarr prototype}
This section describes the methodology used for the AWS parallel Zarr
throughput measurements in Fig~\ref{fig-s3-throughput}
and Fig~\ref{fig-s3-network-throughput}, and timing measurements
detailed in the cloud computing results.
These measurements were run on EC2 instances using jupyter notebooks
hosted on Genomics England's Lifebit CloudOS platform. VCF Zarr chunks
were hosted in a standard profile S3 bucket. Both instance and bucket were
in the same availability zone in the eu-west-2 region. Each
of the parallel Python processes used the aioboto3 library (version 13.2.0) to
fetch chunks concurrently from S3. The underlying aiohttp library
limits the concurrent fetching to 100 requests per process. For decompressing
and computing af-dist the same architecture was used but as fetched chunks
became available in memory in each process, they were decompressed and
computed sequentially. Decompression was performed by the blosc library
without the use of zarr.
